---
title: "2024 UPGG Bootcamp - Data Exploration"
author: "RP Pornmongkolsuk & Natalie Dzikowski"
date: "`r Sys.Date()`"
output: html_document
---

Learning objectives:
* Download and inspect data using R
* Explore data in a scientically motivated way
* Organize and manipulate data in preparation for summary and visualization


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Exploration and Clean-Up

## Refresher
Here's a refresher on how to download the dataset and load it into the "R environment".



## Download data
```{r download data}
library(tidyverse)
periodically_expressed_genes_file <- "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE81nnn/GSE81932/suppl/GSE81932_Dataset02.txt.gz"
periodically_expressed_genes <- read_lines(periodically_expressed_genes_file)
```

## Explore data
Try "printing" the values contained in the variable `periodically_expressed_genes`.
```{r}
periodically_expressed_genes
print(periodically_expressed_genes)
```

`periodically_expressed_genes` is a "vector" data type. It is similar to an "array" or a "list" in some other programming languages, BUT R has its own version of "list." Essentially, R has both "vector" and "list" data types, and they function differently. "Vectors" are what you would think of when you think of Python's "array" or "list"


### Vector
A "vector" is a list of values that, importantly, are of the same type. This is the most common data type in R. 

```{r}
v1 <- c() # Empty vector
v2 <- c(1) # Vector of length 1 and of type numeric
v3 <- c("Hello", "World") # Vector of length 2 and of type character
v4 <- c(1, "hi", TRUE, 1+2) # R is smart enough to coerce the type of all element into one that can be applied to all of them -- in this case, character, designated by " marks.
print(v4)
```

### List
List is a more flexible data tye - it can contain a mixture elements of various data types.



### Data Frame 
This is arguably the most useful feature of R. There's a reason we use spreadsheets. 
And it is particularly efficient too. 